---
title: "Homework 8 R markdown"
author: "Josh Jarvey"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_height: 4
    fig_width: 4.5
  pdf_document:
    fig_height: 4
    fig_width: 4.5
  word_document:
    fig_height: 4
    fig_width: 4.5
---


```{r, setup, include=FALSE}
require(mosaic)   # Load additional packages here 

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

#### <span style="color:Blue">**Intellectual Property:**</span>  
These problems are the intellectual property of the instructors and may not be reproduced outside of this course.

#### <span style="color:Crimson">**Due Date:**</span>  


***  
***  


##########################################################################
## Problem 1: Using Support Vectors for Classification
##########################################################################

<span style="color:DarkViolet">In this problem, you will use a support vector classifier to categorize different species of oak trees to different regions (California vs. the Atlantic coast).  
**Data Set**: Download the *Oak_log.csv* data set and load it into R.  </span>

```{r}
oak = read.csv("C:/Users/joshj/Documents/DS740-Data Mining and Machine Learning/Homework/Week 9/Oak_log.csv")
```

#####################################
### <span style="color:DarkViolet">Question 1</span> **<span style="color:Crimson">(3 points)</span>**:
#####################################

<span style="color:DarkViolet">Make a scatterplot that shows the region, standardized log(acorn size), and standardized log(range) all in the same plot.  Include a legend.  
Enter your R code below.  </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
library(ggformula)

  #create a scatterplot.
gf_point(scale(logSize)~scale(logRange), color = ~Region, data = oak, title = "Standardized acorn size vs. standardized range by Region")
```


#####################################
### <span style="color:DarkViolet">Question 2</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Comment on the appropriateness of a support vector classifier to categorize the regions of species of oak trees. </span>


While there are some overlaps between the classes, a support vector classifier uses a "soft margin" therefore it's ok for a few points to be on the wrong side of the margin or misclassified (which will possibly happen here).


<span style="color:green">**Text Answer**: </span>


#####################################
### <span style="color:DarkViolet">Question 3</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Use a support vector classifier (a SVM with a linear kernel) to categorize the data points.  Use a cost of 1.  
Enter your R code below. </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
library(e1071)

  #fit the svc using cost 1.
svcfit = svm(Region ~ logSize + logRange,
             data = oak, 
             cost = 1, 
             kernel = "linear", 
             type = "C-classification")
```


#####################################
### <span style="color:DarkViolet">Question 4</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Treating the entire data set as the training data, how many Atlantic species are incorrectly classified as California species?</span>


2

<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**:  
```{r}
  #display the confusion matrix. predictions are on the rows, actuals on the columns.
  #there are 2 species the model said california, but were actually atlantic.
table(svcfit$fitted,oak$Region)

```

#####################################
### <span style="color:DarkViolet">Question 5</span> **<span style="color:Crimson">(1 point)</span>**:
#####################################

<span style="color:DarkViolet">How many California species are incorrectly classified as Atlantic species?</span>

<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**:  


0


#####################################
### <span style="color:DarkViolet">Question 6</span> **<span style="color:Crimson">(3 points)</span>**:
#####################################

<span style="color:DarkViolet">Plot the optimal hyperplane (line) with its margins on the same plot from the first question in this problem.  Upload your plot to the *Homework 8: Optimal Hyperplane* discussion board.   </span>

<span style="color:green">**Graph Answer**  </span>: 
  (post to discussion board on D2L)
```{r,echo=FALSE}
  #plot the scaled versions of logSize vs. logRange.
  #use the circle to denote atlantic, and square to denote california - these are the 2 regions.
  #color the atlantic region red, and california region blue.
plot(scale(oak$logRange), scale(oak$logSize),
     main = "Std. acorn size vs. Std. range by region, with classifer margin",
     pch = c(21,22)[oak$Region], 
     col = c("red","blue")[oak$Region])

  #add a legend in the topleft based on the region.
legend("topleft", 
       legend = c("Atlantic", "Califorina"), 
       pch = c(21, 22), 
       col = c("red","blue"))


  #this is beta0, the intercept
b = svcfit$rho 

  #this gives a vector of the coefficients, b1, b2, ...bp. In this instance its b1, and b2.
w = colSums(svcfit$coefs[,1] * svcfit$SV)


#abline(): This function is an easy way to add lines onto a plot.
  # a=the intercept 
  # b=the slope of the line. 


  #this is the line for the decision boundary.
      # beta0 / beta 2.
      # -beta1 / beta2.
abline(b/w[2], -w[1]/w[2])

  #these 2 lines are the margins of the svc.
  #same thing as above, only we add/subtract 1 for the margins. 
abline((b+1)/w[2], -w[1]/w[2], lty=2)
abline((b-1)/w[2], -w[1]/w[2], lty=2)
```



***
***

##########################################################################
## Problem 2:  Using Support Vector Machines for Classification
##########################################################################

<span style="color:DarkViolet">In this problem, you will use support vector machines with different kernels to classify the gas mileage of cars.  
**Data Set**: Load the **Auto** data set, which is in the **ISLR** library.</span>

```{r}
library(ISLR)
data(Auto)

```


#####################################
### <span style="color:DarkViolet">Question 7</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Create a binary variable that equals 1 for cars with gas mileage above the median and a 0 for cars with gas mileage below the median.  Tell R to treat it as a factor.  
Tell R to treat the origin variable as a factor.  
Enter your R code below. </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
  #if mpg > median mpg, =1, else =0.
Auto$goodMPG = as.factor(ifelse(Auto$mpg > median(Auto$mpg), "1","0"))

  #set origin as multi-class factor.
Auto$origin = as.factor(Auto$origin)
```

#####################################
### <span style="color:DarkViolet">Question 8</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Use a support vector classifier to model the binary variable gas mileage as a function of the other variables in Auto.
 
* <span style="color:DarkViolet"> Do not include the continuous variable gas mileage.</span>
* <span style="color:DarkViolet"> Do not include the car’s name.</span>
* <span style="color:DarkViolet"> Use a cost of 1.</span> 

<span style="color:DarkViolet"> Enter your R code below.  </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
  #fitting a support vector classifier.
svcfit = svm(goodMPG~.-mpg-name,
             data = Auto,
             kernel = "linear",
             cost = 1,
             type = "C-classification")

```



#####################################
### <span style="color:DarkViolet">Question 9</span> **<span style="color:Crimson">(3 points)</span>**:
#####################################

<span style="color:DarkViolet">Set your random seed equal to 9, using:
   set.seed(9, sample.kind = "Rounding").  Use a built-in function to perform cross-validation to compare different values of cost:  .001, .01, .1, 1, 5, 10, 100.  
Enter your R code below.  </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
set.seed(9, sample.kind = "Rounding")


  #perform CV10 using various cost values.
tuned.scvfit = tune(svm, 
                    goodMPG~.-mpg-name, 
                    data = Auto, 
                    kernel = "linear", 
                    ranges=(list(cost = c(.001, .01, .1, 1, 5, 10, 100))), 
                    type = "C-classification")

```

#####################################
### <span style="color:DarkViolet">Question 10</span> **<span style="color:Crimson">(1 point)</span>**:
#####################################

<span style="color:DarkViolet">Which cost gave the lowest cross-validation error? </span>


COST = 10

<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**: 

```{r}
  #checking the values of the CV svc.
summary(tuned.scvfit)

```



#####################################
### <span style="color:DarkViolet">Question 11</span> **<span style="color:Crimson">(1 point)</span>**:
#####################################

<span style="color:DarkViolet">What was the cross-validation error of the best model?</span>


<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**: 
  
0.08179487 

***

#####################################
### <span style="color:DarkViolet">Question 12</span> **<span style="color:Crimson">(2 points)</span>**:
#####################################

<span style="color:DarkViolet">Perform cross-validation to compare different values of cost and gamma for a radial model. Use the same values of cost as listed previously: .001, .01, .1, 1, 5, 10, 100. Use gamma =0.5, 1, 2, 3, and 4.  
Enter your R code below.  </span>


<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}
  #perform CV10 with an SVM using various cost and gamma values.
tuned.svmfit = tune(svm, 
                    goodMPG~.-mpg-name, 
                    data = Auto, 
                    kernel = "radial", 
                    ranges=(list(cost = c(.001, .01, .1, 1, 5, 10, 100),
                                 gamma = c(0.5,1,2,3,4))), 
                    type = "C-classification")
```


#####################################
### <span style="color:DarkViolet">Questions 13-14 </span> **<span style="color:Crimson">(2 points, 1 each)</span>**:
#####################################


```{r}
summary(tuned.svmfit)
```

<span style="color:DarkViolet">Which combination of parameters gave the lowest cross-validation error?  </span>

<span style="color:DarkViolet">Cost:  </span>  
<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**:  


COST = 1

<span style="color:DarkViolet">Gamma: </span>  
<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**:  


GAMMA = 3 (actually 1?)

#####################################
### <span style="color:DarkViolet">Question 15 </span> **<span style="color:Crimson">(1 point)</span>**:
#####################################

<span style="color:DarkViolet">What was the cross-validation error of the best model? </span>  

<span style="color:green">**Numeric Answer**  </span> 
**<span style="color:red">(AUTOGRADED on D2L)</span>**:  


0.07384615 (actually 0.0611)

#####################################
### <span style="color:DarkViolet">Question 16</span> **<span style="color:Crimson">(3 points)</span>**:
#####################################

<span style="color:DarkViolet">Use the best model to predict whether the following car would have high or low gas mileage:  
**1977 Chrysler Sunbeam**  
**Cylinders**: 4  
**Engine displacement**:  132.5 cubic inches  
**Horsepower**:  155  
**Weight**:  2,910 lbs  
**Acceleration**:  8.3 seconds  
**Origin**:  American (1)  
Enter your R code below.  </span>

<span style="color:green">**Code Answer**: </span>
```{r, echo=TRUE}

  #pulling out the best model from the CV tuning process.
bestModel = tuned.svmfit$best.model

  #creating a new car's worth of data
    #make sure to match variable types per the original dataframe. Even the amount of factor levels!
newCar = data.frame(
  mpg = 0,
  cylinders = 4,
  displacement = 132.5,
  horsepower = 155,
  weight = 2910,
  acceleration = 8.3,
  year = 1977,
  origin = factor(1, levels = c(1,2,3)),
  name = factor("Chrysler Sunbeam",levels = c(1:304))
  )

  #make a prediction if the car gets good gas mileage.
predict(bestModel,newCar)


```


#####################################
### <span style="color:DarkViolet">Question 17</span> **<span style="color:Crimson">(1 point)</span>**:
#####################################

<span style="color:DarkViolet">What is your prediction?</span>


<span style="color:green">**Multiple choice Answer** </span>
  **<span style="color:red">(AUTOGRADED on D2L)</span>**:  one of  
Low, <<----- correct  
High  
