---
title: "Untitled"
author: "Josh Jarvey"
date: "11/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.0 - Introduction & Data Prep

In this section we begin with reading in the data set, and preparing the relevant variables for the analysis.

We start by reading in the dataset.

```{r}
churn = read.csv("C:/Users/joshj/Documents/DS740-Data Mining and Machine Learning/Homework/Week 13 - Final/Churn.csv")
```

Next, we remove variables that serve no predictive purpose (index, customer id, name) for this analysis. 

```{r}
churn = churn[,-c(1,2,3)]
```

We also need to change the data types of some of the variables to factors. 

```{r}
churn$HasCrCard = as.factor(churn$HasCrCard)
churn$IsActiveMember = as.factor(churn$IsActiveMember)
churn$Exited = as.factor(churn$Exited)
```

Finally, we can check the summary stats of each data type. We're specifically looking for any missing values or outliers. 

```{r}
summary(churn)
```

There are no missing values in this data set, so we can move forward onto the next step of the analysis.

## Data Exploration

In this section we visually explore our dataset - specifically, we're looking at the distributions of the quantitative variables, and also if there are any relationships between variables. 

```{r, fig.width=20, fig.height=10}
library(ggformula)
library(ggpubr)


#gf_boxplot(CreditScore~Exited, data = churn, color = ~Exited)
#gf_boxplot(Age~Exited, data = churn, color = ~Exited)
#gf_boxplot(Balance~Geography, data = churn, color = ~Geography)

scatter = ggarrange(ncol = 2, nrow = 3,
          gf_point(EstimatedSalary~CreditScore, data = churn, color = ~Exited),
          gf_point(EstimatedSalary~Balance, data = churn, color = ~Exited),
          gf_point(EstimatedSalary~Age, data = churn, color = ~Exited),
          gf_point(EstimatedSalary~Tenure, data = churn, color = ~Exited))


histo = ggarrange(ncol = 2, nrow = 3,
          gf_histogram(~EstimatedSalary, data = churn, color = ~Exited, fill = ~Exited),
          gf_histogram(~Balance, data = churn, color = ~Exited, fill = ~Exited),
          gf_histogram(~CreditScore, data = churn, color = ~Exited, fill = ~Exited),
          gf_histogram(~Age, data = churn, color = ~Exited, fill = ~Exited),
          gf_histogram(~Tenure,data = churn, color= ~Exited, fill = ~Exited))


ggarrange(ncol = 2, nrow = 1, histo,scatter)
plot(churn)
```


```{r}
hist(log(churn$Balance+1))
```



## Fitting the model

```{r}
library(nnet)


churn.std = cbind(churn[,c(2,3,8,9,11)],scale(churn[,c(1,4,5,6,7,10)]))

fit1 = nnet(Exited~., data = churn.std, size = 5, decay = 5, maxit = 1000)
```

```{r, fig.width=20, fig.height=10}
library(NeuralNetTools)

plotnet(fit1)



garson(fit1)
```

```{r}
predclass = ifelse(fit1$fitted.values>0.50,1,0)

confusion = table(predclass,churn.std$Exited);confusion

sum(diag(confusion))/sum(confusion)
```


